---
title: "Microarray data GSE26868 (2): Analyse fold changes in gene expression"
subtitle: "BDiB Session 10-March-2022"
author: "Axel Nohturfft"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: "hide"
    css: "source/stylesheet.css"
---

## Overview  
In the **second part** of this workshop we will analyse a table of fold changes in gene expression that was generated with the Geo2r tool on GEO.  

For this analysis we assume that you have used the Geo2r tool on the [GSE26868 page](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE26868) to compare TNF-treated versus control ***THP-1 cells***. The table, including log2FC and pValue data should have been saved as: `"GSE26868_geo2r_results.txt"`.  


## Summary of analysis steps    

1. Clear memory and console
2. Set working directory and load data file
3. Have a brief look at the data frame
4. How many different gene symbols are represented in our data frame?
5. How many different Affy _**probes**_ are represented in our data frame?
6. Clean-up   (i): Keep only relevant columns of the data frame
7. Clean-up  (ii): Remove rows without gene symbol
8. Clean-up (iii): Remove rows with ambiguous gene assignments
9. How many probes per gene?
10. Filter genes by adjusted p.value and fold change
11. Sort data frames by log(fold change)
12. What are the most strongly _**induced**_ genes?
13. What are the most strongly _**suppressed**_ genes?
14. Save symbols of the up and down-regulated genes in text files


## Script setup  
In the first code chunks of this script we will:  

* Clear memory and console  
* Check that all necessary packages are installed  
* Load a few packages  

```{r Clear-memory}
rm(list=ls(all=T)) # Clear memory
cat("\014") # Clear console
options(stringsAsFactors = FALSE)
```

```{r Load-packages, warning=FALSE, message=FALSE}
source("source/packages.R")
```


## Data from GEO2R  
```{r Define-data-file}
gse.id <- "GSE26868"
data.file <- "GSE26868_geo2r_results.txt.gz"
```


Load data file and print the first six rows:  
```{r Read-data}
stopifnot(file.exists(data.file))
df <- read.table(file=data.file, header=TRUE, sep="\t", stringsAsFactors = FALSE)
# The 'head' command prints the first 6 rows of a data frame (or first 6 item of a vector).
# The 'tail' command ...
head(df) %>% 
  knitr::kable(.) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped"))
```

Print number of rows and columns:  
```{r Print-rows-cols, results="asis"}
cat("Number of columns (samples): <b>", ncol(df), "</b><br>\n",
    "Number of rows (probes): <b>", scales::comma(nrow(df)), "</b>",
    sep="")
```

Print column headers:  
```{r Column-headers}
names(df) %>% cat(., sep="\n")
```

How many different Affymetrix probes are represented in our data frame?  
```{r Number-of-probes, class.output="op-asis"}
# print(length(unique(df$ID)))
df$ID %>% as.character %>% unique %>% length %>% print
```

**Student exercise**: How many different gene symbols?  
```{r Exercise1}

```

## CLEAN UP DATA  
### Clean-up (i): Keep only relevant columns of the data frame  
To check results, print first six rows:  
```{r Print-two-rows}
df <- df[,c("ID", "Gene.symbol", "adj.P.Val", "logFC", "Gene.title")]
head(df) %>% 
  knitr::kable(.) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped"))
```

### Clean-up (ii): Remove rows without gene symbol:  
```{r Filter-no-GS, results="asis", class.output="op-asis"}
df2 <- subset(df, Gene.symbol != "") %>% set_rownames(NULL)
cat("Remaining rows: <b>", scales::comma(nrow(df2)), "</b><br>\n",
    "(Number of rows removed: <b>", scales::comma(nrow(df) - nrow(df2)), "</b>)",
    sep="")
```

### Clean-up (iii): Remove rows with ambiguous gene assignments:  
Gene symbols separated by three forward slashes (///) indicate ambiguous gene assignments.  
Print an example:
```{r Example-ambiguous-GS}
df2[72,]
```

We'll use the `grep()` function to search for items in the 'Gene.symbol' column that contain this '///' text. The first argument to `grep()` is the text we are looking for (///), the second argument is a character vector in which to search (the 'Gene.symbol' column in our data frame). grep returns a vector of numbers - these numbers represent the position (index) of where in the vector the search text was found. Type '?grep' at the command prompt or use the RStudio help tab to learn more about grep and related functions.  

```{r Grep-slashes, results="asis"}
index.slashes <- grep("///", df2$Gene.symbol)
# What does the index.slashes vector look like?
# head(index.slashes) # 72 141 176 212 227 312
# How many gene symbols entries are affected?
number.of.bad.symbols <- length(index.slashes) 
cat(paste("Number of rows with 'bad' Gene symbol rows:<b>", number.of.bad.symbols), "</b>")
# 933
```

We can then use this index to subset our data frame and remove the 'bad' rows. (Note the use of the minus sign):
```{r Remove-ambiguous, results="asis"}
df3 <- df2[-index.slashes,]
cat("Remaining rows:<b>", scales::comma(nrow(df3)), "</b>")
```

##  FILTERING MOST AFFECTED GENES    
To identify the most relevant genes in our dataset, we will filter the data based on:  

(i) a maximum adjusted p value and  
(ii) a minimum log~2~(fold change).

### Filtering by adjusted p value:  
The R `subset()` function is very useful for filtering data frames:  
```{r Define-cutoff-p, results="asis"}
adj.P.Val.max <- 0.05
cat("Maximum adjusted p-value:<b>", adj.P.Val.max, "</b>")
```


```{r Filter-by-p, results="asis"}
df4 <- subset(df3, adj.P.Val <= adj.P.Val.max)
cat("Remaining rows: <b>", scales::comma(nrow(df4)), "</b><br>\n",
    "(Number of rows removed: <b>", scales::comma(nrow(df3) - nrow(df4)), "</b>)",
    sep="")
```


### Filtering by log~2~(fold change):  
```{r Define-cutoff-FC, results="asis"}
logFC.min <- 1
cat("Minimum log2 fold change:<b>", logFC.min, "</b>")
```

```{r Filter-by-FC, results="asis"}
df5 <- subset(df4, abs(logFC) >= logFC.min)
cat("Remaining rows: <b>", scales::comma(nrow(df5)), "</b><br>\n",
    "(Number of rows removed: <b>",
    scales::comma(nrow(df4) - nrow(df5)),
    "</b>)", sep="")
```


### Print the whole filtered data frame:  
```{r Print-filtered-df}
DT::datatable(df5, rownames=FALSE)
```


### SELECT UP AND DOWNREGULATED GENES
We can now extract the up and down-regulated genes and save them for further analysis.  

#### Up-regulated genes:  
```{r def-up, results="asis", message=FALSE}
df.up <- df5 %>% 
  # Keep only rows where fold change is positive:
  dplyr::filter(logFC > 0) %>% 
  # Sort decreasing by fold change:
  dplyr::arrange(desc(logFC)) %>% 
  # Remove row names:
  set_rownames(NULL)
cat("Number of up-regulated probes:<b>", nrow(df.up), "</b>")
```


Print first the resulting data frame with up-regulated probes only:  
```{r Print-df-up}
DT::datatable(df.up, rownames = FALSE)
```


#### Down-regulated genes:  
```{r df-down, results="asis"}
df.down <- df5 %>% 
  # Keep only rows where fold change is negative:
  dplyr::filter(logFC < 0) %>% 
  # Sort increasing by fold change:
  dplyr::arrange(logFC) %>% 
  # Remove row names:
  set_rownames(NULL)
cat("Number of down-regulated probes:<b>", nrow(df.down), "</b>")
```

Print the resulting data frame with down-regulated probes only:  
```{r Print-df-down}
DT::datatable(df.down, rownames = FALSE)
```


#### Save gene symbols:  
```{r gs-up, results="asis"}
up <- df.up$Gene.symbol %>% unique %>% sort
details::details(paste(up, collapse=", "), lang=NULL, summary="Show upregulated genes")
```

Save gene symbols:
```{r Save-gs-up, results="asis"}
out.file.up <-  paste0(gse.id, "_up.txt")
cat(up, file=out.file.up, sep="\n")
cat("File saved:<b>", out.file.up, "</b>")
```

```{r gs-down, results="asis"}
down <- df.down$Gene.symbol %>% unique %>% sort
details::details(paste(down, collapse=", "), lang=NULL, summary="Show downregulated genes")
```

Save gene symbols:
```{r Save-gs-down, results="asis"}
out.file.down <- paste0(gse.id, "_down.txt")
cat(down, file=out.file.down, sep="\n")
cat("File saved:<b>", out.file.down, "</b>")
```

#### Show files in working directory:  
```{r Print-files}
cat(list.files(), sep="\n")
```


## NEXT ...  
The genes could now be analyzed e.g. for 'gene set enrichment' to see if many of them fall into a known category.  
Such analyses can be done directly in R, or more conveniently by uploading the gene symbols to websites such as:

* [GOrilla](http://cbl-gorilla.cs.technion.ac.il/): looks for enrichment with 'gene ontology' (GO) categories.
* [enrichr](http://amp.pharm.mssm.edu/Enrichr/) and  
* [GSEA](http://software.broadinstitute.org/gsea/index.jsp) ('gene set enrichment analysis') allow to search for overlap among a larger set of gene categories.

## Session info  
```{r SESSION-INFO-DATE, results="asis"}
cat("Date:", format(Sys.time(), "%a %d-%b-%Y %H:%M:%S"), "<br>\n")
```

```{r SESSION-INFO-PLATFORM}
si <- sessioninfo::session_info()
si.platform.df <- data.frame(
  setting = names(si$platform),
  value   = unlist(si$platform),
  stringsAsFactors = FALSE,
  row.names = NULL
)
DT::datatable(si.platform.df,
              options = list(columnDefs = list(list(className = 'dt-left', targets = 0:1)),
                             dom = 't'),
              rownames = FALSE,
              class="stripe")
```

```{r SESSION-INFO-PACKAGES}
si$packages %>% 
  as.data.frame() %>% 
  dplyr::select(package, version=loadedversion, date, source) %>% 
  DT::datatable(options = list(columnDefs = list(list(className = 'dt-left', targets = 0:3))),
              rownames = FALSE,
              class="stripe")
  
```




